---
title: "Application of Central Limit Theorem to Exponential Distribution - Example"
author: "Amade A."
date: "December 2, 2015"
output: pdf_document
---
## Overview

This paper describes the application of Central Limit Theorem to the distribution of means of exponential distribution. The emphasis is put on empirical presentation of a problem, with a random sample of 1000 means and comparing their distribution with the theoretical distribution of the mean and theoretical variance.


## Simulations

```{r, echo = FALSE}
# Setting the global number formatting option
library(knitr)
knit_hooks$set(inline = function(x) {
  prettyNum(x, big.mark=",")
})
```

```{r , echo = TRUE}
n <- 1000
sampleSize <- 40
lambda <- 0.2
```

For the purpose of presentation, random sample of **`r n`** means of exponential distribution needs to be generated. Each mean is calculated based on **`r sampleSize`** values from exponential distribution with parameter $\lambda$ of **`r lambda`**. In turn, in order to obtain the sample, **`r (n*sampleSize)`** random numbers need to be generated.

The following code generates the sample.

```{R, echo = TRUE}
set.seed(2015)
means <- apply(matrix(rexp(n*sampleSize, rate = lambda),n,sampleSize),1,mean)
```

## Sample Mean versus Theoretical Mean

The sample mean can be calculated with the following code.

```{r, echo = TRUE}
sampleMean <- mean(means)
sampleMean
```

Theoretical mean of a sample coming from exponential distribution equals $\frac{1}{\lambda}$.
Assuming $\lambda$ of `r lambda`, theoretical mean is $\frac{1}{\lambda} = \frac{1}{`r lambda`} = `r 1/lambda`$.

The best way to assess the magnitude of a difference in exmpirical and theoretical means is looking at a graph, which shows the histogram of expirical means.

```{r}
library(ggplot2)
m <- ggplot() + aes(x=means)
m <- m + geom_histogram(fill = "navy", binwidth = .15) 
m <- m + geom_vline(xintercept = (1/lambda), colour = "red", size = 1)
m + ggtitle("Theoretical mean vs. empirical mean distribution")
```

Theoretical sample mean is shown as red line. It can be clearly seen on the graph, that the means are centered around its value (`r 1/lambda`). In relative terms, sample mean is only `r abs(round(((sampleMean - 1/lambda)/(1/lambda))*100,2))`% different from theoretical mean (calculated as $abs(\frac{sample \hspace{2pt} mean - theoretical \hspace{2pt} mean}{theoretical \hspace{2pt} mean})$).

## Sample Variance versus Theoretical Variance

Theoretical variance of the sampling distribution of the mean is calculated as:

$\sigma_{M}^{2} = \frac{\sigma^{2}}{N}$

where $\sigma^{2}$ is a theoretical variance of the distribution and $N$ is a sample size. In the case of exponential distribution, theoretical variance equals $\frac{1}{\lambda^{2}}$. With $\lambda = `r lambda`$ and $N = `r sampleSize`$, theoretical variance of the sampling distribution of the mean equals 
```{r, echo = TRUE}
theorVar <- (1/lambda^2)/(sampleSize)
theorVar
```

Unbiased estimator of the population variance, based on the sample data, calculated using *var()* function equals `r round(var(means),3)`. It is `r round(abs((var(means)-theorVar)/(theorVar)*100),2)`% different from theoretical variance.

Variability of the sample mean is shown on the graph below.

```{r, echo = TRUE}
SDmarker <- function(x) {  
    if (x > 2*sqrt(theorVar)){
        "x > 2 SD"
    } else if (x > sqrt(theorVar)){
        "1 SD < x <= 2 SD"
    } else {
        "x <= 1 SD"
    }
}
SDband <- as.factor(sapply(abs(means - sampleMean),SDmarker))

m <- ggplot() + aes(x=means, fill = SDband)
m <- m + geom_histogram(binwidth = .15) 
m <- m + geom_vline(xintercept = sampleMean, colour = "red", size = .5)
m <- m + geom_vline(xintercept = sampleMean + c(-1,1)*sqrt(theorVar),
                    colour = "green", size = .5)
m <- m + geom_vline(xintercept = sampleMean + c(-2,2)*sqrt(theorVar),
                    colour = "blue", size = .5)
m + ggtitle("Variability of the sample mean")
```

Sample mean (`r sampleMean`) is marked in red. Area marked in pink, represents the means with values within 1 standard deviation (its theoretical value) form the sample mean. Area in green - those between 1 and 2 standard deviations from the mean. Means, which fell more than 2 standard deviations from the mean are represented by the light blue area.

## Distribution

In order to test the assumption of normality of the sample mean distribution, a histogram of empirical values will be compared with a normal distribution with a mean and variance equal theoretical mean and theoretical variance of the analyzed sampling distribution - $N(`r 1/lambda`, `r theorVar`)$. Line representing normal distribution is shown in red.

```{r, echo = TRUE}

m <- ggplot() + aes(x=means)
m <- m + geom_histogram(aes(y = ..density..), binwidth = .15, fill = "navy") 
m <- m + stat_function(fun = dnorm,
                       args = list(mean = 1/lambda, sd = sqrt(theorVar)), colour = "red")
m + ggtitle("Empirical distribution vs. normal distribution")
```

Based on the visual comparison, it can be concluded that the empirical distribution is approximately normal.

## Sources

1. [LaTEX code generator](https://www.codecogs.com/latex/eqneditor.php)